Type-Safe Modular Parsing

### 2016-11-15 paper/sections/ObjectAlgebras.tex
More.

%But we can only abstract over the whole interface type, instead of working on its type parameters, as the number of type parameters is not fixed, where we will discuss more on multiple syntax later. In summary, \lstinline{"|||"} produces a combined parser, which takes a compound algebra as its parameter.

@@

>>>>

@@

\subsection{Parsing with Multiple Syntax}\label{subsec:differentsyntax}

One intuition of using Object Algebras, as illustrated above, is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, or realize different operations on structures, without modifying existing code, meanwhile separate compilation is supported. Another one is that Object Algebras benefit us from using generics, where we can abstract multiple syntax in the type parameters, and define less classes, simplifying code comparatively. Below we refactor the last example in Section~\ref{subsec:parsingwithopen}, by defining a new Object Algebra interface with two cases \lstinline{intT} and \lstinline{lamE}. Meanwhile a parser is implemented for that language.
\begin{lstlisting}
trait LamAlg[E, T] {
    def intT() : T
    def lamE(x : String, t : T, e : E) : E
}

trait LamParser[E, T] {
    type Pair = (Parser[E], Parser[T])
    val pE : LamAlg[E, T] => (=> Pair) => Parser[E] = alg => p =>
        ("\\" ~> ident) ~ (":" ~> p._2) ~ ("." ~> p._1)
            ^^ { case x ~ t ~ e => alg.lamE(x, t, e) }
    val pT : LamAlg[E, T] => (=> Pair) => Parser[T] = alg => p =>
        "Int" ^^ { _ => alg.intT() }
}
\end{lstlisting}
\lstinline{LamParser} contains parsers for both expressions and types. From the code we observe that the multi-sorts of Object Algebra interface
exactly express multiple syntax. At this point, to combine these two small components together we need some ``glue code''. But instead of writing composition by hand, we reimplement the combinator \lstinline{|||}, to compose one parser with type \lstinline{Alg1 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} and another one with type \lstinline{Alg2 =>} \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} using alternative. Note that we can only abstract over the whole algebra interface type, instead of working on its type parameters, as the number of type parameters is not fixed. In summary, \lstinline{"|||"} produces a combined parser, which takes a compound algebra as its parameter:
\begin{lstlisting}
implicit class Combinator[Alg1, Alg2, R, E](x : Alg1 => (=> R) => Parser[E]) {
    def |||(y : Alg2 => (=> R) => Parser[E]): (Alg1 with Alg2) => (=> R) => Parser[E]
        = alg => l => x(alg)(l) ||| y(alg)(l)
}
\end{lstlisting}
Now it is sufficient to merge \lstinline{ExprParser} and \lstinline{LamParser} together:
\begin{lstlisting}
trait ExprLamParser[E, T] {
    val pExpr = new ExprParser[E](){}
    val pLam = new LamParser[E, T](){}
    val pExprE : ExprAlg[E] => (=> pLam.Pair) => Parser[E] =
        alg => p => pExpr.pE(alg)(p._1)

    val pE = pExprE ||| pLam.pE
    val pT = pLam.pT
}
\end{lstlisting}
Parsers for expressions and types are connected using \lstinline{|||}, respectively. Hence the isolation for different syntax
is guaranteed by the type system. The only left problem is that we still need to replace the parser in \lstinline{ExprParser} with \lstinline{pExprE}
by updating the fix-point parameter. In the next section we will address this problem, and meanwhile our implementation enhances modularity in various aspects.

### 2016-11-15 paper/sections/OpenRecursion.tex
More.

abstract class Expr
case class Var(x : String) extends Expr
val pVar : Parser[Expr] = ident ^^ Var
\end{lstlisting}
Note that we are omitting lexing and demo code here. And we use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} throughout the paper. Furthermore, we would like the parser to support applications.
Not only the implementation of that parser is needed, but also its corresponding datatype:
\begin{lstlisting}
case class App(e1 : Expr, e2 : Expr) extends Expr
val pApp : Parser[Expr] = pVar ~ pVar ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Such a parser \lstinline{pApp} can only parse one-layer applications like \lstinline{"x y"}. Hence we revise it using recursion:
\begin{lstlisting}
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp
    p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}
Note that we are composing \lstinline{pVar} and \lstinline{pApp} using alternative. Alternation is a general operation to compose different parsers, since they represent different grammar rules with the logical relation ``or'' between one another. In this example, both sub-expressions of an application can be either a single variable or again an application. Now \lstinline{pApp} manages to parse an arbitrary number of applications like \lstinline{"x y z ..."} in a right-associative way.
Nevertheless, even though we are not considering operations but the creation of objects only, such an approach does not turn out to be a modular extension, because if numeric literals are further accepted in our small language, the grammar becomes:
\begin{lstlisting}
e ::= x | e ' ' e | n
\end{lstlisting}
Besides writing a new parser for literals, we need to rewrite the existing \lstinline{pApp}, since applications can now contain numbers, like \lstinline{"x 3"} and so on.
\begin{lstlisting}
case class Lit(n : Int) extends Expr
val pLit : Parser[Expr] = numericLit ^^ Lit
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp ||| pLit
    p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}
We have observed that \lstinline{pApp} is indeed constructed in a non-modular way, because the \lstinline{p} standing for recursive parser needs to be updated on every extension. Actually in the grammar, the recursive \lstinline{"e"} in the second case should not only include all the existing cases like \lstinline{pApp} and \lstinline{pVar}, but also take future extensions into account (we can have arbitrarily more cases). This is where we are inspired to use open recursion. To define \lstinline{pApp} as an ``open'' function, we add a parameter \lstinline{p} to it, representing the explicit self-reference of our whole parser, open to future extensions. Note that \lstinline{p} should be defined as a by-name parameter using \lstinline{"=>"}.
\begin{lstlisting}
val pApp : Fix[Parser[Expr]] = p => p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Here the type \lstinline{Fix[Parser[Expr]]}, namely \lstinline{(=>} \lstinline{Parser[Expr])} \lstinline{=>} \lstinline{Parser[Expr]} is exactly what we want for a parser. Similarly \lstinline{pVar} is redefined also as a function using the explicit fix-point \lstinline{p}. To be more convenient for composition, we define an auxiliary combinator \lstinline{"|||"} for parsers of type \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} with the same \lstinline{R} and \lstinline{E}. Such a function is overloaded and can be invoked as an infix operator when defined in an implicit class. It is really implemented using the general \lstinline{merge} function in Section~\ref{subsec:openrecursion}.
\begin{lstlisting}
val pVar : Fix[Parser[Expr]] = p => ident ^^ Var

@@

>>>>

abstract class Expr
case class Var(x : String) extends Expr
val pVar : Parser[Expr] = ident ^^ Var
\end{lstlisting}
Note that we are omitting lexing and demo code here. And we use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} throughout the paper. Furthermore, we would like the parser to support applications.
Not only the implementation of that parser is needed, but also its corresponding datatype:
\begin{lstlisting}
case class App(e1 : Expr, e2 : Expr) extends Expr
val pApp : Parser[Expr] = pVar ~ pVar ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Such a parser \lstinline{pApp} can only parse one-layer applications like \lstinline{"x y"}. Hence we revise it using recursion:
\begin{lstlisting}
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp
    p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}
Note that we are composing \lstinline{pVar} and \lstinline{pApp} using alternative. Alternation is a general operation to compose different parsers, since they represent different grammar rules with the logical relation ``or'' between one another. In this example, both sub-expressions of an application can be either a single variable or again an application. Now \lstinline{pApp} manages to parse an arbitrary number of applications like \lstinline{"x y z ..."} in a right-associative way.
Nevertheless, even though we are not considering operations but the creation of objects only, such an approach does not turn out to be a modular extension, because if numeric literals are further accepted in our small language, the grammar becomes:
\begin{lstlisting}
e ::= x | e ' ' e | n
\end{lstlisting}
Besides writing a new parser for literals, we need to rewrite the existing \lstinline{pApp}, since applications can now contain numbers, like \lstinline{"x 3"} and so on.
\begin{lstlisting}
case class Lit(n : Int) extends Expr
val pLit : Parser[Expr] = numericLit ^^ Lit
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp ||| pLit
    p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}
We have observed that \lstinline{pApp} is indeed constructed in a non-modular way, because the \lstinline{p} standing for recursive parser needs to be updated on every extension. Actually in the grammar, the recursive \lstinline{"e"} in the second case should not only include all the existing cases like \lstinline{pApp} and \lstinline{pVar}, but also take future extensions into account (we can have arbitrarily more cases). This is where we are inspired to use open recursion. To define \lstinline{pApp} as an ``open'' function, we add a parameter \lstinline{p} to it, representing the explicit self-reference of our whole parser, open to future extensions. Note that \lstinline{p} should be defined as a by-name parameter using \lstinline{"=>"}.
\begin{lstlisting}
val pApp : Fix[Parser[Expr]] = p => p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Here the type \lstinline{Fix[Parser[Expr]]}, namely \lstinline{(=>} \lstinline{Parser[Expr])} \lstinline{=>} \lstinline{Parser[Expr]} is exactly what we want for a parser. Similarly \lstinline{pVar} is redefined also as a function using the explicit fix-point \lstinline{p}. To be more convenient for composition, we define an auxiliary combinator \lstinline{"|||"} for parsers of type \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} with the same \lstinline{R} and \lstinline{E}. It is more general using \lstinline{R} instead of \lstinline{Parser[E]}, because there might be multiple fix-points in the parameter. Such a function is overloaded and can be invoked as an infix operator when defined in an implicit class. It is implemented using the general \lstinline{merge} function in Section~\ref{subsec:openrecursion}.
\begin{lstlisting}
val pVar : Fix[Parser[Expr]] = p => ident ^^ Var

@@

Another weakness is that the only little extensibility is just fragile when it comes to grammars with multiple different syntax. For instance, in the beginning we only have variables and applications in \lstinline{Expr}, to have types in our language, the \lstinline{Int} type and lambda expressions with annotation are added. See the new grammar below:
\begin{lstlisting}
e ::= x | e ' ' e | \x:t.e
t ::= "Int"
\end{lstlisting}

We proceed to implement those parsers. An issue is that we need both a \lstinline{Parser[Expr]} and a \lstinline{Parser[Type]} in the fixpoint.
Hence we use a pair in the parameter:
\begin{lstlisting}
type Pair = (Parser[Expr], Parser[Type])

abstract class Type
case class IntT extends Type
val pInt : (=> Pair) => Parser[Type] =
    p => "Int" ^^ { _ => IntT() }

case class Lam(x : String, t : Type, e : Expr) extends Expr
val pLam : (=> Pair) => Parser[Expr] =
    p => ("\\" ~> ident) ~ (":" ~> p._2) ~ ("." ~> p._1) ^^
        { case x ~ t ~ e => Lam(x, t, e) }

val pVar2 : (=> Pair) => Parser[Expr] = p => pVar(p._1)
val pApp2 : (=> Pair) => Parser[Expr] = p => pApp(p._1)

val pET : Fix[Pair] = {
    val pE = pVar2 ||| pApp2 ||| pLam
    val pT = pInt
    p => (pE(p), pT(p))
}
\end{lstlisting}
Based on the old \lstinline{pVar} and \lstinline{pApp}, we implement \lstinline{pVar2} and \lstinline{pApp2} which take the pair as their fix-points instead.
In that case the parsers are consistent, and hence can be composed using the combinator \lstinline{|||}.
The last value \lstinline{pET} integrates the four single parsers defined before, where parsers for expressions and parsers for types are composed respectively. Furthermore, we use \lstinline{fix(p)} to obtain the pair of two parsers we want: one for expressions, the other for types. The parser \lstinline{fix(p)._1} parsers expressions like \lstinline{"\x:Int.x x"} successfully. Yet meanwhile, some drawbacks of this approach come into sight. With different syntax, it
is necessary to extend the fix-point to contain more parsers, in which case usually some structures like tuples are used,
making code tedious and less elegant. Moreover, we need to update such a fix-point in the arguments of all previously defined parsers, so as to make parsers consistent for the combinator (things get even worse without that combinator). It introduces more boilerplate code, which highly affects code reuse.
